---
title: "TriMet Service Area Desire Lines"
author: "M. Edward (Ed) Borasky"
date: '2018-06-06'
output: html_notebook
bibliography: desire_lines.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lodes)
library(tidyverse)
library(stplanr)
library(sp)
library(tmap)
library(sf)
```

## Introduction
As part of the Hack Oregon 2018 season, TriMet gave the Transportation Systems team a dataset summarizing ridership metrics by service stop, totaled over three-month quarters. The ridership dataset covers March 2001 through November 2017, and includes breakdowns by weekday, Saturday and Sunday service.

To provide context for the weekday service, which includes commuters, we analyzed another dataset, the US Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) [@LEHD2018] LEHD Origin-Destination Employment Statistics (LODES) data [@LODES2018]. This dataset covers the years 2002 through 2015.

The LODES dataset provides job counts by the workers' home and work census blocks, where the blocks are defined by the 2010 decennial census. We use the `lodes` package {@Rudis2017] to acquire the LODES data and `stplanr` [Lovelace2018a] to compute the desire lines.

## Acquiring the LODES data

### Downloading the crosswalk file
First, we download the "crosswalk" file for Oregon. This file contains the geographic information for the census blocks.
```{r message=FALSE}
or_xwalk <- read_xwalk("or", download_dir = "~/Raw")
or_xwalk$tabblk2010 <- as.character(or_xwalk$tabblk2010) 
```
### Downloading the origin-destination data
```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
or_od_main_JT00 <- tibble()
for (ixyear in 2002:2015) {
  temp <-   read_lodes("or", "od", "main", "JT00", ixyear, download_dir = "~/Raw")
  temp$w_geocode <- as.character(temp$w_geocode)
  temp$h_geocode <- as.character(temp$h_geocode)
  temp <- temp %>% mutate(year = ixyear) %>% select(-createdate)
  or_od_main_JT00 <- or_od_main_JT00 %>% bind_rows(temp)
  rm(temp); gc()
}
```

### Workplace analysis
First, we want to look at the geographical properties of workplaces. To do this, we create an aggregated geotagged data frame [@Pebesma2005, @Bivand2013]
```{r}
## sum all the job counts by year and workplace census block (w_geocode)
or_workplaces <- or_od_main_JT00 %>% 
  group_by(year, w_geocode) %>%
  summarize_at(.funs = sum, .vars = vars(S000:SI03)) %>% 
  left_join(or_xwalk, by = c("w_geocode" = "tabblk2010")) %>% 
  select(
    year, w_geocode, S000:SI03,
    cbsaname, stplcname,
    blklondd, blklatdd
  ) %>% 
  ungroup()

## convert to a "simple features" data frame
or_workplaces <- SpatialPointsDataFrame(
  coords = or_workplaces %>% select(blklondd, blklatdd),
  data = or_workplaces,
  proj4string = CRS("+init=epsg:4269")
) %>% st_as_sf()
```

### Make the "zones" data frame
The process `stplanr` uses to compute the desire lines has two inputs, the _flow_ and the _zones_ [@Lovelace2018a, vignette "Introducing stplanr"]. The `zones` data frame is a SpatialPointsDataFrame as defined by the `sp` package [@Pebesma2005]
```{r}
zones <- SpatialPointsDataFrame(
  coords = select(or_xwalk, blklondd, blklatdd),
    data = select(or_xwalk, tabblk2010, blklondd, blklatdd),
    proj4string = CRS("+init=epsg:4269"))
```

## Computing the desire lines

### Metro 2002
We start by isolating the top 20 Oregon workplaces in the Portland metro area. We use `S000` (all jobs) as the jobs count.
```{r}
top_metro_workplaces_2002 <- or_workplaces %>% filter(
  cbsaname == "Portland-Vancouver-Hillsboro, OR-WA",
  year == 2002
) %>% 
  mutate(jobs = S000) %>% arrange(desc(jobs)) %>% top_n(20, jobs)
```

Now compute the flow - number of commuters into the workplaces by home and work census blocks. To keep things simple we only take the top 50 entries.
```{r}
flow_metro_2002 <- or_od_main_JT00 %>% filter(
  year == 2002,
  w_geocode != h_geocode) %>% 
  semi_join(top_metro_workplaces_2002, by = "w_geocode") %>% 
  select(w_geocode, h_geocode, jobs = S000) %>% 
  arrange(desc(jobs)) %>% top_n(50, jobs)
```

Compute the desire lines with `od2line` from `stplanr` [@Lovelace2018a].
```{r}
desire_lines_metro_2002 <- od2line(flow_metro_2002, zones)
```

Make an interactive map with `qtm` from `tmap` [@Tennekes2018]. This is a Leaflet map. The "+" and "-" buttons can be used to zoom in and out and you can drag the map around with the mouse. You can select the basemap and whether to display the desire lines and workplace circles with the "layers" button below the "-". If you click on one of the circles you'll see the number of jobs.

Note that the bulk of the top 20 workplaces / desire lines are on the west side. The three leftmost points are Intel in Hillsboro. The cluster of three points in Beaverton are Nike and Tektronix. The dark orange (9728 jobs) circle near the Willamette is the OHSU complex, and the circle near the 217 - 26 interchange (4368 jobs) is Providence St. Vincent's. 
```{r}
tmap_mode("view")
qtm(
  desire_lines_metro_2002, 
  lines.lwd = "jobs", 
  lines.scale = 5, 
  lines.col = "firebrick"
) + qtm(
  top_metro_workplaces_2002, symbols.scale = 1, symbols.col = "jobs")
save_tmap(filename = "desire_lines_metro_2002.html")
```

### Metro 2015
Let's repeat the process for the last year in our dataset, 2015.
```{r}
top_metro_workplaces_2015 <- or_workplaces %>% filter(
  cbsaname == "Portland-Vancouver-Hillsboro, OR-WA",
  year == 2015
) %>% 
  mutate(jobs = S000) %>% arrange(desc(jobs)) %>% top_n(20, jobs)

flow_metro_2015 <- or_od_main_JT00 %>% filter(
  year == 2015,
  w_geocode != h_geocode) %>% 
  semi_join(top_metro_workplaces_2015, by = "w_geocode") %>% 
  select(w_geocode, h_geocode, jobs = S000) %>% 
  arrange(desc(jobs)) %>% top_n(50, jobs)

desire_lines_metro_2015 <- od2line(flow_metro_2015, zones)
```

The dominance of the west side in 2015 is even more pronounced.
```{r}
tmap_mode("view")
qtm(
  desire_lines_metro_2015, 
  lines.lwd = "jobs", 
  lines.scale = 5, 
  lines.col = "firebrick"
) + qtm(
  top_metro_workplaces_2015, symbols.scale = 1, symbols.col = "jobs")
save_tmap(filename = "desire_lines_metro_2015.html")

```

### City of Portland (minus OHSU) 2015
To see the flows with the city of Portland, we have to remove the census block containing the OHSU complex, `410510058002014`.
```{r}
top_portland_workplaces_2015 <- or_workplaces %>% filter(
  stplcname == "Portland city, OR",
  year == 2015,
  w_geocode != "410510058002014"
) %>% 
  mutate(jobs = S000) %>% arrange(desc(jobs)) %>% top_n(20, jobs)

flow_portland_2015 <- or_od_main_JT00 %>% filter(
  year == 2015,
  w_geocode != h_geocode) %>% 
  semi_join(top_portland_workplaces_2015, by = "w_geocode") %>% 
  select(w_geocode, h_geocode, jobs = S000) %>% 
  arrange(desc(jobs)) %>% top_n(50, jobs)

desire_lines_portland_2015 <- od2line(flow_portland_2015, zones)
```

With the focus on the city of Portland with OHSU removed, you can see the diversity of the commuters' transit needs.
```{r}
tmap_mode("view")
qtm(
  desire_lines_portland_2015, 
  lines.lwd = "jobs", 
  lines.scale = 5, 
  lines.col = "firebrick"
) + qtm(
  top_portland_workplaces_2015, symbols.scale = 1, symbols.col = "jobs")
save_tmap(filename = "desire_lines_portland_2015.html")
```

## Session info
```{r}
devtools::session_info()
```
## References
