---
title: "TriMet Service Area Desire Lines"
author: "M. Edward (Ed) Borasky"
date: '2018-06-09'
output: html_notebook
bibliography: desire_lines.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lodes)
library(tidyverse)
library(stplanr)
library(sp)
library(tmap)
library(sf)

```

## Introduction

### About the data
As part of the Hack Oregon 2018 season, TriMet gave the Transportation Systems team a dataset summarizing ridership metrics by service stop, totaled over three-month quarters. The ridership dataset covers March 2001 through November 2017, and includes breakdowns by weekday, Saturday and Sunday service.

To provide context for the weekday service, which includes commuters, we analyzed another dataset, the US Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) [@LEHD2018] LEHD Origin-Destination Employment Statistics (LODES) data [@LODES2018a]. This dataset covers the years 2002 through 2015.

The LODES dataset provides job counts by the workers' home and work census blocks, where the blocks are defined by the 2010 decennial census [@LODES2018b]. 

### About the software
We use the `lodes` package [@Rudis2017] to acquire the LODES data, `stplanr` [@Lovelace2018a] to compute the desire lines, `tmap` [@Tennekes2018] to generate the interactive maps, both `sp` [@Pebesma2005] and `sf` [@Bivand2013] for spatial processing, and the `tidyverse` [@Wickham2017] for data wrangling.

## Acquiring the LODES data

### Download the crosswalk files
First, we download the "crosswalk" files for Oregon and Washington. These files contain the geographic information for the census blocks. We filter down to only the Portland - Vancouver - Hillsboro core-based statistical area (CBSA).
```{r message=FALSE}
xwalk <- bind_rows(
  read_xwalk("or", download_dir = "~/Raw/LODES") %>% 
    filter(cbsaname == "Portland-Vancouver-Hillsboro, OR-WA") %>% 
    select(-createdate),
  read_xwalk("wa", download_dir = "~/Raw/LODES") %>% 
    filter(cbsaname == "Portland-Vancouver-Hillsboro, OR-WA") %>% 
    select(-createdate)
)
xwalk$tabblk2010 <- as.character(xwalk$tabblk2010)
gc()

```

### Make the "zones" SpatialPointsDataFrame
The process `stplanr` uses to compute the desire lines has two inputs, the _flow_ and the _zones_ [@Lovelace2018a, vignette "Introducing stplanr"]. The `zones` data frame is a SpatialPointsDataFrame, as defined by the `sp` package [@Pebesma2005], that provides the coordinates of the interior points of the census blocks.

Note that the Census Bureau uses the coordinate reference system (CRS) EPSG:4269, but our tools require EPSG:4326. So we re-project to EPSG:4326.
```{r}
to_4326 <- function(sp_object) {
  spTransform(sp_object, CRSobj = CRS("+init=epsg:4326"))
}

zones <- SpatialPointsDataFrame(
  coords = select(xwalk, blklondd, blklatdd),
    data = select(xwalk, tabblk2010, blklondd, blklatdd),
    proj4string = CRS("+init=epsg:4269"))
zones <- to_4326(zones)

```

### Make the `xwalk_4_joins` data frame
The `xwalk` data frame contains numerous fields we won't be using, so we filter down to just the ones we need, creating the `xwalk_4_joins` data frame.
```{r}
xwalk_4_joins <- xwalk %>% 
  select(tabblk2010, cbsaname, stplcname, blklatdd, blklondd)

```

### Download the origin-destination data
The origin-destination data is split into two segments, "main" and "aux". "Main" covers jobs where the workers live in the state; "aux" covers jobs where the workers live outside the state. Because some Oregon workers live in Washington and vice versa, we acquire both "main" and "aux" for Oregon and Washington.
```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
# define a function to fetch a file
fetch_od <- function(state, segment, ixyear, xwalk_4_joins) {
  temp <-read_lodes(
    state, "od", segment, "JT00", ixyear, download_dir = "~/Raw/LODES")
  temp$w_geocode <- as.character(temp$w_geocode)
  temp$h_geocode <- as.character(temp$h_geocode)
  temp <- temp %>% select(-createdate)
  temp <- temp %>% mutate(year = ixyear)
  temp <- temp %>% 
    semi_join(xwalk_4_joins, by = c("w_geocode" = "tabblk2010"))
  temp <- temp %>% 
    semi_join(xwalk_4_joins, by = c("h_geocode" = "tabblk2010")) 
  gc()
  return(temp)
}

metro_od <- tibble()
for (ixyear in 2002:2015) {
  for (state in c("or", "wa")) {
    for (type in c("main", "aux")) {
      temp <- fetch_od(state, type, ixyear, xwalk_4_joins)
      metro_od <- metro_od %>% bind_rows(temp)
      rm(temp); gc()
    }
  }
}

```

### Download the residence area characteristics data
Because we have workers who live and work in both Washington and Oregon, we download both states' residence and workplace area characteristics.
```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
fetch_rac <- function(state, ixyear, xwalk_4_joins) {
  temp <-read_lodes(
    state, "rac", "S000", "JT00", ixyear, download_dir = "~/Raw/LODES")
  temp$h_geocode <- as.character(temp$h_geocode)
  temp <- temp %>% select(-createdate)
  temp <- temp %>% mutate(year = ixyear) 
  temp <- temp %>% 
    inner_join(xwalk_4_joins, by = c("h_geocode" = "tabblk2010"))
  gc()
  return(temp)
}

metro_rac <- tibble()
for (ixyear in 2002:2015) {
  for (state in c("or", "wa")) {
    temp <- fetch_rac(state, ixyear, xwalk_4_joins)
    metro_rac <- metro_rac %>% bind_rows(temp)
    rm(temp); gc()
  }
}

metro_rac <- SpatialPointsDataFrame(
  coords = select(metro_rac, blklondd, blklatdd),
  data = metro_rac,
  proj4string = CRS("+init=epsg:4269")
)
gc()
metro_rac <- to_4326(metro_rac)
gc()
metro_rac <- metro_rac %>% as("sf")
gc()

```

### Plot top 50 residence areas
We make interactive maps with `qtm` from `tmap` [@Tennekes2018]. This is a Leaflet map. The "+" and "-" buttons can be used to zoom in and out and you can drag the map around with the mouse to center it before zooming in. You can select the basemap and whether to display plotted features with the "layers" button below the "-". If you click on one of the circles you'll see the number of workers.

```{r}
tmap_mode("view")
qtm(
  metro_rac %>% 
    filter(year == 2015) %>% 
    mutate(Employees = C000) %>% 
    arrange(desc(Employees)) %>% 
    top_n(50, Employees),
  symbols.col = "Employees",
  symbols.size = "Employees"
)
save_tmap(filename = "metro_rac_2015.html")

```

### Download the workplace area characteristics data
```{r message=FALSE, warning=FALSE, error=FALSE, results='hide'}
fetch_wac <- function(state, ixyear, xwalk_4_joins) {
  temp <-read_lodes(
    state, "wac", "S000", "JT00", ixyear, download_dir = "~/Raw/LODES")
  temp$w_geocode <- as.character(temp$w_geocode)
  temp <- temp %>% select(-createdate)
  temp <- temp %>% mutate(year = ixyear) 
  temp <- temp %>% 
    inner_join(xwalk_4_joins, by = c("w_geocode" = "tabblk2010"))
  gc()
  return(temp)
}

metro_wac <- tibble()
for (ixyear in 2002:2015) {
  for (state in c("or", "wa")) {
    temp <- fetch_wac(state, ixyear, xwalk_4_joins)
    metro_wac <- metro_wac %>% bind_rows(temp)
    rm(temp); gc()
  }
}

metro_wac <- SpatialPointsDataFrame(
  coords = select(metro_wac, blklondd, blklatdd),
  data = metro_wac,
  proj4string = CRS("+init=epsg:4269")
)
gc()
metro_wac <- to_4326(metro_wac)
gc()
metro_wac <- metro_wac %>% as("sf")
gc()

```

## Plot the workplace map
This map shows the dominance of the West Side for jobs. The large circle near the Willamette is the OHSU complex on Marquam Hill. On the MAX Blue Line, the two circles in Hillsboro and Orenco are Intel, the 9,205-worker circle in Beaverton is Nike, and the 4,611-worker circle near the Sunset Highway - 217 interchange is the St. Vincent's complex.
```{r}
tmap_mode("view")
qtm(
  metro_wac %>% 
    filter(year == 2015) %>% 
    mutate(Employees = C000) %>% 
    arrange(desc(Employees)) %>% 
    top_n(50, Employees),
  symbols.col = "Employees",
  symbols.size = "Employees"
)
save_tmap(filename = "metro_wac_2015.html")

```

## Computing the desire lines

### Desire line computation functions
To compute the desire lines, `stplanr` needs two datasets - the `flow` and the `zones`. The zones don't change - they're simply a table of the spatial locations of the census blocks, and we've already computed them.

But there are two ways we can look at the flows - from the point of view of the workers or the point of view of the employers. Either way, it's a three-step process:

1. Pick the top residence or workplace census blocks. This will be our point of view.
2. Compute the flows out of the blocks for residences or into the blocks for workplaces.
3. Compute the desire lines.

This next code chunk defines the functions we'll use.
```{r}
top_residences <- function(rac, ixyear, rows) {
  residences <- rac %>% 
    filter(year == ixyear) %>% 
    mutate(Workers = C000) %>% 
    select(-C000:-CS02) %>% 
    arrange(desc(Workers)) %>% 
    top_n(rows, Workers)
}

residence_flow <- function(od, ixyear, top_residences, rows) {
  flow <- od %>% 
  filter(
    year == ixyear,
    w_geocode != h_geocode
  ) %>% 
  semi_join(top_residences, by = "h_geocode") %>% 
  select(w_geocode, h_geocode, year, Jobs = S000) %>% 
  arrange(desc(Jobs)) %>% 
  top_n(rows, Jobs)
}

top_workplaces <- function(wac, ixyear, rows) {
  workplaces <- wac %>% 
    filter(year == ixyear) %>% 
    mutate(Employees = C000) %>% 
    select(-C000:-CFS05) %>% 
    arrange(desc(Employees)) %>% 
    top_n(rows, Employees)
}

workplace_flow <- function(od, ixyear, top_workplaces, rows) {
  flow <- od %>% 
  filter(
    year == ixyear,
    w_geocode != h_geocode
  ) %>% 
  semi_join(top_workplaces, by = "w_geocode") %>% 
  select(w_geocode, h_geocode, year, Jobs = S000) %>% 
  arrange(desc(Jobs)) %>% 
  top_n(rows, Jobs)
}

desire_lines <- function(flow, zones) {
  lines <- od2line(flow, zones) %>% as("sf")
}

# default number of residence or workplace census blocks to use
npoints <- 20

# default number of flow lines to compute
nlines <- 100

```

### Workers' viewpoint for 2002
Let's look at the desire lines from the workers' point of view for 2002 first.
```{r}
ixyear <- 2002
top_r <- top_residences(metro_rac, ixyear, npoints)
flow_r <- residence_flow(metro_od, ixyear, top_r, nlines)
desire_r <- desire_lines(flow_r, zones)
```

### Plot the desire lines
This map shows the desire lines from the point of view of the workers - where should the service stops and routes be to get them to work and back home. The desire lines flow outward from residence census blocks.
```{r}
tmap_mode("view")
qtm(
  desire_r, 
  lines.lwd = "Jobs", 
  lines.scale = 10, 
  lines.col = "firebrick"
) + 
qtm(
  top_r, 
  symbols.scale = 2, 
  symbols.size = "Workers", 
  symbols.col = "Workers"
)
save_tmap(filename = "residence_desire_lines_2002.html")
```

### 2002 from the employers' point of view
```{r}
top_w <- top_workplaces(metro_wac, ixyear, npoints)
flow_w <- workplace_flow(metro_od, ixyear, top_w, nlines)
desire_w <- desire_lines(flow_w, zones)

tmap_mode("view")
qtm(
  desire_w, 
  lines.lwd = "Jobs", 
  lines.scale = 10, 
  lines.col = "firebrick"
) + 
qtm(
  top_w, 
  symbols.scale = 2, 
  symbols.size = "Employees", 
  symbols.col = "Employees"
)
save_tmap(filename = "workplace_desire_lines_2002.html")

```

## Tabulate desire lines for all years
To provide a convenient dataset for future plotting, we create two `sf` data frames with the two variants of desire lines.
```{r}
# start with the 2002 data
residence_desire_lines <- desire_r
workplace_desire_lines <- desire_w

for (ixyear in 2003:2015) {
  top_r <- top_residences(metro_rac, ixyear, npoints)
  top_w <- top_workplaces(metro_wac, ixyear, npoints)
  flow_r <- residence_flow(metro_od, ixyear, top_r, nlines)
  flow_w <- workplace_flow(metro_od, ixyear, top_w, nlines)
  desire_r <- desire_lines(flow_r, zones)
  desire_w <- desire_lines(flow_w, zones)
  residence_desire_lines <- residence_desire_lines %>% rbind(desire_r)
  workplace_desire_lines <- workplace_desire_lines %>% rbind(desire_w)
}
```

## Session info
```{r}
devtools::session_info()
```
## References
