---
title: "TriMet Service Area Desire Lines"
author: "M. Edward (Ed) Borasky"
date: "2018-06-06"
output: github_document
bibliography: desire_lines.bib
always_allow_html: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install `lodes` from GitHub if needed
if (!require(lodes)) {
  devtools::install_github("hrbrmstr/lodes", force = TRUE)
  library(lodes)
}
library(tidyverse)
library(stplanr)
library(sp)
library(tmap)
```
## Introduction
As part of the Hack Oregon 2018 season, TriMet gave the Transportation Systems team a dataset summarizing ridership metrics by service stop, totaled over three-month quarters. This dataset covers March 2001 through November 2017, and includes breakdowns for weekday, Saturday and Sunday service.

To provide context for the weekday service, which includes commuters, we analyzed another dataset, the US Census Bureau's Longitudinal Employer-Household Dynamics (LEHD) LEHD Origin-Destination Employment Statistics (LODES) data [@LODES2018]. This dataset covers the years 2002 through 2015.

The LODES dataset provides job counts by the workers' home and work census blocks, where the blocks are defined by the 2010 decennial census. We use the `lodes` package {@Rudis2017] to acquire the LODES data and `stplanr` [Lovelace2018a] to compute the desire lines.

## Acquiring the LODES data

### Downloading the crosswalk file
First, we download the "crosswalk" file for Oregon. This file contains the geographic information for the census blocks.
```{r}
or_xwalk <- read_xwalk("or", download_dir = "~/Raw")
or_xwalk$tabblk2010 <- as.character(or_xwalk$tabblk2010) 
```

### Filtering down to the TriMet service area
The TriMet service area is contained in three Oregon counties - Clackamas, Multnomah and Washington. So we filter the crosswalk table down to just those three counties.
```{r}
or_xwalk <- or_xwalk %>% filter(
  ctyname == "Clackamas County, OR" |
    ctyname == "Multnomah County, OR" |
    ctyname == "Washington County, OR"
)
```

### Acquring the origin-destination files
```{r}
or_od_main_JT00_2015 <- 
  read_lodes("or", "od", "main", "JT00", 2015, download_dir = "~/Raw")

or_od_main_JT00_2015$w_geocode <- as.character(or_od_main_JT00_2015$w_geocode)
or_od_main_JT00_2015$h_geocode <- as.character(or_od_main_JT00_2015$h_geocode)
```

### Filter down to the TriMet service area
```{r}
or_od_main_JT00_2015 <- or_od_main_JT00_2015 %>% 
  semi_join(or_xwalk, by = c("w_geocode" = "tabblk2010")) %>% 
  semi_join(or_xwalk, by = c("h_geocode" = "tabblk2010"))
```

## Computing the desire lines

### Computing the flow
```{r}
flow <- or_od_main_JT00_2015 %>%  
  select(w_geocode, h_geocode, S000) %>% 
  arrange(desc(S000)) %>% top_n(200, S000)
```

### Making the "zones" file
The process `stplanr` uses to compute the desire lines has two inputs, the _flow_ and the _zones_ [@Lovelace2018a, vignette "Introducing stplanr"]. The `zones` data frame is a SpatialPointsDataFrame as defined by the `sp` package [@Pebesma2005, @Bivand2013]
```{r}
zones <- SpatialPointsDataFrame(
  coords = select(or_xwalk, blklondd, blklatdd),
    data = select(or_xwalk, tabblk2010, blklondd, blklatdd),
    proj4string = CRS("+init=epsg:4269"))
```

### Make the desire lines and home points
```{r}
home_points <- flow %>% select(h_geocode) %>% unique() %>% 
  left_join(or_xwalk, by = c("h_geocode" = "tabblk2010")) %>% 
  select(h_geocode, blklondd, blklatdd)
home_points <- SpatialPointsDataFrame(
  coords = select(home_points, blklondd, blklatdd),
    data = select(home_points, h_geocode, blklondd, blklatdd),
    proj4string = CRS("+init=epsg:4269")
)
  
desire_lines <- od2line(flow, zones)
```

```{r}
tmap_mode("view")
qtm(desire_lines)

```

## Session info
```{r}
devtools::session_info()
```
## References
